---
title: "Basic Probability Concepts"
format: 
  html:
    code-fold: true
---

**Events:** In probability theory, an event is an outcome or a set of outcomes of a random experiment. Events are the events we are interested in, and they can range from simple outcomes to complex combinations.

**Sample Space:** The sample space is the set of all possible outcomes of a random experiment. It encompasses every possible scenario, providing a comprehensive view of the potential results.

**Example 1:** Consider rolling a six-sided die.

```{r}
# Events and Sample Space Example

# Define the sample space for rolling a six-sided die
sample_space <- 1:6

# Define the event "rolling an even number"
event_A <- c(2, 4, 6)

# Define the event "rolling a number greater than 4"
event_B <- c(5, 6)

# Display the events and sample space
cat("Event A (Rolling an even number):", event_A, "\n")
cat("Event B (Rolling a number greater than 4):", event_B, "\n")
cat("Sample Space:", sample_space, "\n")
```

**Example 2:** Consider flipping a coin with Head and Tail

```{r}
# Sample Space Example 2

# Define the sample space for flipping a coin
coin_sample_space <- c("Heads", "Tails")

# Display the sample space for flipping a coin
cat("Sample Space for Flipping a Coin:", coin_sample_space, "\n")

```

## Probability of an Event

The probability of an event is a measure of the likelihood that the event will occur. It is expressed as a value between 0 and 1, where 0 indicates impossibility, 1 indicates certainty, and values in between represent varying degrees of likelihood.

#### Independence and Mutually Exclusive Events

**Independence:** Events A and B are independent if the occurrence of one does not affect the occurrence of the other.

**Mutually Exclusive (Disjoint):** Events A and B are mutually exclusive if they cannot both occur at the same time.

```{r}
# Probability of an Event Example

# Number of favorable outcomes for event A
favorable_outcomes_A <- length(event_A)

# Total number of possible outcomes
total_outcomes <- length(sample_space)

# Calculate the probability of event A
probability_A <- favorable_outcomes_A / total_outcomes

# Display the probability of event A
cat("Probability of Event A (Rolling an even number):", probability_A, "\n")

```

The result of a coin flip using a Bernoulli distribution in R, you can create a bar plot that represents the probability of getting Heads and Tails. The bar heights will correspond to the probabilities of success (Heads) and failure (Tails). 
```{r}
# Set the probability of getting Heads
p_heads <- 0.5

# Calculate the probability of getting Tails
p_tails <- 1 - p_heads

# Create a bar plot
barplot(c(p_heads, p_tails), names.arg = c("Heads", "Tails"),
        col = c("blue", "red"), ylim = c(0, 1),
        main = "Coin Flip Probability",
        xlab = "Outcome", ylab = "Probability")

# Add probability labels to the bars
text(1:2, c(p_heads, p_tails), labels = c(p_heads, p_tails), pos = 3, col = "black")

```
```{r}

```
**Example 3:** Here is an example where we use a Gaussian distribution to model continuous data, such as the distribution of heights in a population. In this case, we'll use the 'dnorm' function to generate a Gaussian distribution and create a histogram to visualize it.
```{r}
# Set the parameters for the Gaussian distribution
mean_height <- 170  # mean height in centimeters
sd_height <- 10      # standard deviation of height

# Generate a sample of heights from a Gaussian distribution
num_samples <- 1000
heights <- rnorm(num_samples, mean = mean_height, sd = sd_height)

# Create a histogram to visualize the Gaussian distribution
hist(heights, breaks = 30, col = "skyblue", main = "Gaussian Distribution of Heights",
     xlab = "Height (cm)", ylab = "Frequency")

# Add a curve representing the theoretical Gaussian distribution
curve(dnorm(x, mean = mean_height, sd = sd_height), col = "darkblue", lwd = 2, add = TRUE)


```

