---
title: "Random Variables and Probability Distributions"
format: 
  html:
    code-fold: true
---

In probability theory, a **random variable** is a variable whose possible values are outcomes of a random phenomenon. Understanding random variables is crucial in modeling uncertainty and variability in real-world scenarios. Each random variable is associated with a **probability distribution**, which describes the likelihood of different outcomes.

#### **Types of Random Variables**

**Discrete Random Variables**

-   Examples: Number of heads in coin flips, count of customers arriving at a store.

-   Rationale: Takes on a countable number of distinct values.

**Continuous Random Variables**

-   Examples: Height of individuals, temperature, stock prices.

-   Rationale: Takes on an uncountable infinite number of possible values.

#### **Probability Distributions**

**Probability Mass Function (PMF)**

1.  For discrete random variables.

2.  Gives the probability of each possible outcome.

```{r}
# Example PMF for a six-sided die
outcomes <- 1:6
probabilities <- rep(1/6, 6)
pmf <- data.frame(Outcome = outcomes, Probability = probabilities)

# Display the pmf
print(pmf)
```
To visualize this discrete distribution, let's create a bar plot
```{r}
barplot(probabilities, names.arg = outcomes, col = "lightblue",
        main = "Probability Mass Function for a Six-Sided Die",
        xlab = "Outcome", ylab = "Probability")
```

**Probability Density Function (PDF)**

1.  For continuous random variables.

2.  Describes the likelihood of a range of values.

```{r}
# Example PDF for a standard normal distribution
x <- seq(-3, 3, length.out = 100)
pdf <- data.frame(X = x, Probability = dnorm(x))

# Display the pdf
print(pdf)

```
We  can visualize the PDF using a line plot as shown below

```{r}
# Plot the PDF
plot(pdf$X, pdf$Probability, type = "l", col = "blue",
     main = "Probability Density Function for Standard Normal Distribution",
     xlab = "X", ylab = "Probability Density")

# Add a vertical line at x = 0 for the mean
abline(v = 0, col = "red", lty = 2)

```

#### **Expectation and Variance**

-   **Expectation (Mean):** Represents the average value of a random variable.

```{r}
# Example calculation of expectation for a discrete random variable
mean_discrete <- sum(outcomes * probabilities)

# Display the mean
cat("mean:", mean_discrete, "\n")
```

-   **Variance:** Measures the spread or dispersion of a random variable.

```{r}
# Example calculation of variance for a discrete random variable
variance_discrete <- sum((outcomes - mean_discrete)^2 * probabilities)

# Display the variance
cat("variance:", variance_discrete, "\n")
```

## 
