---
title: "Basic Probability Concepts"
format: 
  html:
    code-fold: true
---

**Events:** In probability theory, an event is an outcome or a set of outcomes of a random experiment. Events are the events we are interested in, and they can range from simple outcomes to complex combinations.

**Sample Space:** The sample space is the set of all possible outcomes of a random experiment. It encompasses every possible scenario, providing a comprehensive view of the potential results.

**Example 1:** Consider rolling a six-sided die.

```{r}
# Events and Sample Space Example

# Define the sample space for rolling a six-sided die
sample_space <- 1:6

# Define the event "rolling an even number"
event_A <- c(2, 4, 6)

# Define the event "rolling a number greater than 4"
event_B <- c(5, 6)

# Display the events and sample space
cat("Event A (Rolling an even number):", event_A, "\n")
cat("Event B (Rolling a number greater than 4):", event_B, "\n")
cat("Sample Space:", sample_space, "\n")
```

**Example 2:** Consider flipping a coin with Head and Tail

```{r}
# Sample Space Example 2

# Define the sample space for flipping a coin
coin_sample_space <- c("Heads", "Tails")

# Display the sample space for flipping a coin
cat("Sample Space for Flipping a Coin:", coin_sample_space, "\n")

```

## Probability of an Event

The probability of an event is a measure of the likelihood that the event will occur. It is expressed as a value between 0 and 1, where 0 indicates impossibility, 1 indicates certainty, and values in between represent varying degrees of likelihood.

#### Independence and Mutually Exclusive Events

**Independence:** Events A and B are independent if the occurrence of one does not affect the occurrence of the other.

**Mutually Exclusive (Disjoint):** Events A and B are mutually exclusive if they cannot both occur at the same time.

```{r}
# Probability of an Event Example

# Number of favorable outcomes for event A
favorable_outcomes_A <- length(event_A)

# Total number of possible outcomes
total_outcomes <- length(sample_space)

# Calculate the probability of event A
probability_A <- favorable_outcomes_A / total_outcomes

# Display the probability of event A
cat("Probability of Event A (Rolling an even number):", probability_A, "\n")

```

## 
